---
title: "Diagnostic test evaluation with perfect reference test"
author: "Paolo Eusebi"
date: "16/09/2021"

theme: metropolis
aspectratio: 43
colortheme: seahorse

output:
  beamer_presentation: 
      slide_level: 2
---

```{r setup, message=F, echo=F, warning=F, render=F}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
```

## Four phases in architecture of diagnostic research

- **Phase I** Do test results in patients with the target disorder differ from those in normal people? 

- **Phase II** Are patients with certain test results more likely to have the target disorder than patients with other test results?

- **Phase III** Does the test result distinguish patients with and without the target disorder among patients in whom it is clinically reasonable to suspect that the disease is present?

- **Phase IV** Do patients who undergo this diagnostic test fare better (in their ultimate health outcomes) than similar patients who are not tested?


## Measures of diagnostic accuracy

```{r, echo=FALSE}
df <- tibble("  " = c("T+", "T-"),
             "D+" = c("TP", "FN"),
             "D-" = c("FP", "TN"))

df %>% kable()
```


## Measures of diagnostic accuracy

```{r,echo=FALSE}
df <- tibble("  " = c("T+", "T-"),
             "D+" = c("TP", "FN"),
             "D-" = c("FP", "TN"))

df %>% kable()
```

- Sensitivity = TP/D+
- Specificity = TN/D-
- PPV = TP/T+
- NPV = TN/T-

## Measures of diagnostic accuracy

- Sensitivity and specificity do not depend on the disease prevalence.

- PPV and NPV depend on the sensitivity, specificity, and the disease prevalence.

$PPV=\frac{Se \cdot p}{Se \cdot p + (1-Sp) \cdot (1-p)}$
  
$NPV=\frac{Sp \cdot (1-p)}{(1-Se) \cdot p + Sp \cdot (1-p)}$


## Measures of diagnostic accuracy

- Frequencies
```{r, echo=FALSE}

df <- tibble("  " = c("T+", "T-"),
             "D+" = c("y[1]", "y[2]"),
             "D-" = c("y[3]", "y[4]"))

df %>% kable()
```

- Probabilities
```{r, echo=FALSE}

df <- tibble("  " = c("T+", "T-"),
             "D+" = c("prob[1]", "prob[2]"),
             "D-" = c("prob[3]", "prob[4]"))
df %>% kable()
```

- Prevalence = prob[1]+prob[2]


## Bayesian model

```{r}
"model {

# likelihood
  y[1:4] ~ dmulti(prob[1:4], n)

  prob[1] <- p * Se
  prob[2] <- p * (1 - Se)
  prob[3] <- (1 - p) * (1 - Sp)
  prob[4] <- (1 - p) * Sp

# priors
  p ~ dbeta(1, 1)
  Se ~ dbeta(1,1)
  Sp ~ dbeta(1,1)
"
```

## Bayesian model

Let's code!
