---
title: "Diagnostic test evaluation with perfect reference test"
author: "Paolo Eusebi"
date: "16/09/2021"

theme: metropolis
aspectratio: 43
colortheme: seahorse

output:
  beamer_presentation: 
      slide_level: 2
---

```{r setup, message=F, echo=F, warning=F, render=F}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
```

## Four phases in architecture of diagnostic research

- **Phase I** Do test results in patients with the target disorder differ from those in normal people? 

- **Phase II** Are patients with certain test results more likely to have the target disorder than patients with other test results?

- **Phase III** Does the test result distinguish patients with and without the target disorder among patients in whom it is clinically reasonable to suspect that the disease is present?

- **Phase IV** Do patients who undergo this diagnostic test fare better (in their ultimate health outcomes) than similar patients who are not tested?


## Measures of diagnostic accuracy

```{r, echo=FALSE}
df <- tibble("  " = c("T+", "T-"),
             "D+" = c("TP", "FN"),
             "D-" = c("FP", "TN"))

df %>% kable()
```

- TP = True Positives
- FP = False Positives
- TN = True Negatives
- FN = False Negatives


## Measures of diagnostic accuracy

```{r,echo=FALSE}
df <- tibble("  " = c("T+", "T-"),
             "D+" = c("TP", "FN"),
             "D-" = c("FP", "TN"))

df %>% kable()
```

- Sensitivity = TP/D+
- Specificity = TN/D-
- PPV = TP/T+
- NPV = TN/T-

## Measures of diagnostic accuracy

- Sensitivity and specificity do not depend on the disease prevalence.

- PPV and NPV depend on the sensitivity, specificity, and the disease prevalence.

$$PPV=\frac{Se \cdot p}{Se \cdot p + (1-Sp) \cdot (1-p)}$$
  
$$NPV=\frac{Sp \cdot (1-p)}{(1-Se) \cdot p + Sp \cdot (1-p)}$$


## Measures of diagnostic accuracy: Bayesian reasoning

Suppose a new HIV test has 95% Se and 98% Sp and is to be used in a population with a HIV prevalence of 1/1000.

We can simulate the diagnostic results in a population of 100,000 individuals.

```{r,echo=FALSE}
df <- tibble("  " = c("T+", "T-"),
             "HIV+" = c("95", "5"),
             "HIV-" = c("1,998", "97,902"))

df %>% kable()
```

- PPV = 95/(95+1,998)= 0.0045 (Only 4.5%!)


Example from Spiegelhalter et al. *Bayesian approaches to clinical trials and health-care evaluation* John Wiley & Sons, 2004.


## Measures of diagnostic accuracy: Bayesian reasoning

$$P(\theta|Y) = \frac{P(\theta)\times P(Y|\theta)}{P(Y)}$$

$$\theta=\theta_0 (HIV+),\theta_1 (HIV-)$$

- Data probability $$P(Y)=P(\theta_0) \cdot P(Y|\theta_0)+P(\theta_1) \cdot P(Y|\theta_1) = $$ 
$$= 0.001 \cdot 0.95+0.999 \cdot 0.02 = 0.02093$$

- Prior $$P(\theta_0) = 1/1000 = 0.001$$ 

- Likelihood  (95% Se) $$P(Y|\theta_0) = 0.95$$

- Posterior $$P(\theta_0|Y) = (0.001 \cdot 0.95)/0.02093 = 0.0045$$


## Measures of diagnostic accuracy

- Frequencies
```{r, echo=FALSE}

df <- tibble("  " = c("T+", "T-"),
             "D+" = c("y[1]", "y[2]"),
             "D-" = c("y[3]", "y[4]"))

df %>% kable()
```

- Probabilities
```{r, echo=FALSE}

df <- tibble("  " = c("T+", "T-"),
             "D+" = c("prob[1]", "prob[2]"),
             "D-" = c("prob[3]", "prob[4]"))
df %>% kable()
```


## Measures of diagnostic accuracy

```{r, echo=FALSE}

df <- tibble("  " = c("T+", "T-"),
             "D+" = c("prob[1]", "prob[2]"),
             "D-" = c("prob[3]", "prob[4]"))
df %>% kable()
```

We can define parameters as function of the cell probabilities: 

- p = prob[1]+prob[2]

- Se = prob[1]/(prob[1]+prob[2])

- Sp = prob[4]/(prob[3]+prob[4])

Or cell probabilities in terms of parameters:   

- prob[1] = Se*p

- prob[2] = (1-Se)*p

- prob[3] = Sp*(1-p)

- prob[4] = Sp*(1-p)


## Bayesian model

```{r}
"model {

# likelihood
  y[1:4] ~ dmulti(prob[1:4], n)

  prob[1] <- p * Se
  prob[2] <- p * (1 - Se)
  prob[3] <- (1 - p) * (1 - Sp)
  prob[4] <- (1 - p) * Sp

# priors
  p ~ dbeta(1, 1)
  Se ~ dbeta(1,1)
  Sp ~ dbeta(1,1)
"
```

## Bayesian model

Let's code!


