---
title: "Diagnostic test evaluation with imperfect reference test"
author: "Paolo Eusebi"
date: "16/09/2021"

theme: metropolis
aspectratio: 43
colortheme: seahorse

output:
  beamer_presentation: 
      slide_level: 2
---

```{r setup, include=TRUE, message=FALSE, warning = FALSE, echo=FALSE, render=FALSE}
knitr::opts_chunk$set(echo = T)
library(tidyverse)
library(kableExtra)
library(runjags)
```


## Recap

- Fitting models using MCMC is easy with JAGS / runjags

- But we must **never forget** to check convergence and effective sample size!

- More complex models become easy to implement

- So how do we extend these models to multiple diagnostic tests?


# Hui-Walter models for 2 tests and 1 population

## Hui-Walter Model

- A particular model formulation that was originally designed for evaluating diagnostic tests in the absence of a gold standard

- Not necessarily (or originally) Bayesian but often implemented using Bayesian MCMC
  
- But evaluating an imperfect test against another imperfect test is a bit like pulling a rabbit out of a hat

- If we don't know the true disease status, how can we estimate sensitivity or specificity for either test?

## Hui-Walter Model: 2 tests, 1 population

- D+
```{r, echo=F}
df <- tibble("  "  = c("T1+", "T1-"),
             "T2+" = c("p*Se1*Se2", "p*(1-Se1)*Se2"),
             "T2-" = c("p*Se1*(1-Se2)", "p*(1-Se1)*(1-Se2)"))
df %>% kable()
```

- D-
```{r, echo=F}
df <- tibble("  "  = c("T1+", "T1-"),
             "T2+" = c("p*Sp1*Sp2", "p*(1-Sp1)*Sp2"),
             "T2-" = c("p*Sp1*(1-Sp2)", "p*(1-Sp1)*(1-Sp2)"))
df %>% kable()
```

- 5 parameters: Se1, Se2, Sp1, Sp2, p


## Hui-Walter Model: 2 tests, 1 population

- We don't know the disease status >> Marginalize over D+ and D-

```{r, echo=F}
df <- tibble("  "  = c("T1+", "T1-"),
             "T2+" = c("p*Se1*Se2+p*Sp1*Sp2", "p*(1-Se1)*Se2+p*(1-Sp1)*Sp2"),
             "T2-" = c("p*Se1*(1-Se2)+p*Sp1*(1-Sp2)", "p*(1-Se1)*(1-Se2)+p*(1-Sp1)*(1-Sp2)"))
df %>% kable()
```

- 5 parameters and 3 equations >> Non-identifiable model!

- We can't fit this model with standard likelihood procedures.

- Let's try do go Bayesian!


## Model Specification

```{r, echo=F}
hw_2t_1p <- c("model{
  y ~ dmulti(prob, TotalTests)
  
  # Test1- Test2-
	prob[1] <- (p*((1-se[1])*(1-se[2]))) + ((1-p)*((sp[1])*(sp[2])))
  # Test1+ Test2-
	prob[2] <- (p*((se[1])*(1-se[2]))) + ((1-p)*((1-sp[1])*(sp[2])))
  # Test1- Test2+
	prob[3] <- (p*((1-se[1])*(se[2]))) + ((1-p)*((sp[1])*(1-sp[2])))
  # Test1+ Test2+
	prob[4] <- (p*((se[1])*(se[2]))) + ((1-p)*((1-sp[1])*(1-sp[2])))
	", " 
  p ~ dbeta(1, 1)
  se[1] ~ dbeta(1, 1)
  sp[1] ~ dbeta(1, 1)
  se[2] ~ dbeta(1, 1)
  sp[2] ~ dbeta(1, 1)
  #data# y, TotalTests
  #monitor# p, prob, se, sp
  #inits# p, se, sp
}
")
cat(hw_2t_1p, sep='', file='hw_2t_1p.bug')
```


```{r comment='', echo=FALSE}
cat(hw_2t_1p[1], sep='\n')
```

## Model Specification

```{r comment='', echo=FALSE}
cat(hw_2t_1p[2], sep='\n')
```

## Data, inititial values, run
```{r, echo=FALSE}
df <- tibble("  " = c("T1+", "T1-"),
             "T2+" = c(48, 12),
             "T2-" = c(4, 36))
df %>% kable()
```


```{r, message=FALSE, warning=FALSE, results='hide'}
y <- c(48, 12, 4, 36)
TotalTests <- sum(y)
p <- list(chain1=0.05, chain2=0.95)
se <- list(chain1=c(0.5,0.99), chain2=c(0.99,0.5))
sp <- list(chain1=c(0.5,0.99), chain2=c(0.99,0.5))
results <- run.jags(hw_2t_1p, n.chains=2)
```

[Remember to check convergence and effective sample size!]

---

```{r, eval=FALSE}
results
```

```{r echo=FALSE}
res <- summary(results)[,c(1:3,9,11)]
res[] <- round(res, 3)
knitr::kable(res)
```

- Note the wide credible intervals!


## Practicalities

- These models need a lot of data and/or strong priors for one of the tests

- Convergence is more problematic than usual

- Be very careful with the order of combinations in dmultinom!

- Check your results carefully to ensure they make sense!


## Label Switching

How to interpret a test with Se=0% and Sp=0%?

The test is perfect - we are just holding it upside down...

We can force Se + Sp >= 1:

```{r eval=FALSE}
  se[1] ~ dbeta(1, 1)
  sp[1] ~ dbeta(1, 1)T(1-se[1], )
```

or:

```{r eval=FALSE}
  se[1] ~ dbeta(1, 1)T(1-sp[1], )
  sp[1] ~ dbeta(1, 1)
```

But not both!

This allows the test to be useless, but not worse than useless


## Simulating data

Analysing simulated data is useful to check that we can recover parameter values.

```{r}
se1 <- 0.9; sp1 <- 0.95
se2 <- 0.8; sp2 <- 0.99
prevalence <- 0.5; N <- 100
truestatus <- rbinom(N, 1, prevalence)
Test1 <- rbinom(N, 1, (truestatus * se1) + 
                  ((1-truestatus) * (1-sp1)))
Test2 <- rbinom(N, 1, (truestatus * se2) + 
                  ((1-truestatus) * (1-sp2)))
t <- table(Test1, Test2)
y <- as.numeric(t)
```

Can we recover these parameter values?
  
## Exercise
  
- Modify the code in the Hui-Walter model to force tests to be no worse than useless

- Simulate data and recover parameters for N=10, 100 or 10000


# Hui-Walter models for 2 tests and N populations

## Hui-Walter : 2 tests, 2 pops

- Population 1

\scriptsize
```{r, echo=F}
df <- tibble("  "  = c("T1+", "T1-"),
             "T2+" = c("p1*Se1*Se2+p1*Sp1*Sp2", "p1*(1-Se1)*Se2+p1*(1-Sp1)*Sp2"),
             "T2-" = c("p1*Se1*(1-Se2)+p1*Sp1*(1-Sp2)", "p1*(1-Se1)*(1-Se2)+p1*(1-Sp1)*(1-Sp2)"))
df %>% kable()
```

- Population 2
```{r, echo=F}
df <- tibble("  "  = c("T1+", "T1-"),
             "T2+" = c("p2*Se1*Se2 + p2*Sp1*Sp2", "p2*(1-Se1)*Se2+p2*(1-Sp1)*Sp2"),
             "T2-" = c("p2*Se1*(1-Se2)+p2*Sp1*(1-Sp2)", "p2*(1-Se1)*(1-Se2)+p2*(1-Sp1)*(1-Sp2)"))
df %>% kable()
```
\normalsize

- 6 parameters and 6 equations >> Identifiable model!



## Hui-Walter models with multiple populations

- Basically an extension of the single-population model

- Works best with multiple populations each with differing prevalence

- Including an unexposed population works well BUT be wary of assumptions regarding constant sensitivity/specificity across populations with very different types of infections


## Independent intercepts for populations

```{r eval=FALSE}
model{
  for(p in 1:Populations){
    Tally[1:4, p] ~ dmulti(prob[1:4, p], TotalTests[p])
    # Test1- Test2- Pop1
    prob[1, p] <- (prev[p] * ((1-se[1])*(1-se[2]))) + ((1-prev[p]) * ((sp[1])*(sp[2])))
    ## etc ##
    prev[p] ~ dbeta(1, 1)
  }
  se[1] ~ dbeta(HPSe[1,1], HPSe[1,2])T(1-sp[1], )
  sp[1] ~ dbeta(HPSp[1,1], HPSp[1,2])
  se[2] ~ dbeta(HPSe[2,1], HPSe[2,2])T(1-sp[2], )
  sp[2] ~ dbeta(HPSp[2,1], HPSp[2,2])
  #data# Tally, TotalTests, Populations, HPSe, HPSp
  #monitor# prev, prob, se, sp
  #inits# prev, se, sp
}
```


## Auto Hui-Walter

We would usually start with individual-level data in a dataframe:
  
```{r}
se1 <- 0.9; se2 <- 0.8; sp1 <- 0.95; sp2 <- 0.99
prevalences <- c(0.1, 0.5, 0.9)
N <- 100
simdata <- data.frame(Population = sample(seq_along(prevalences), N, replace=TRUE))
simdata$probability <- prevalences[simdata$Population]
simdata$truestatus <- rbinom(N, 1, simdata$probability)
simdata$Test1 <- rbinom(N, 1, (simdata$truestatus * se1) + ((1-simdata$truestatus) * (1-sp1)))
simdata$Test2 <- rbinom(N, 1, (simdata$truestatus * se2) + ((1-simdata$truestatus) * (1-sp2)))
```

---
  
```{r, eval=FALSE}
head(simdata)
```

```{r, echo=FALSE}
head(simdata)
```

[Except that probability and truestatus would not normally be known!]

---
  
The model code and data format for an arbitrary number of populations (and tests) can be determined automatically

There is a function (provided in the GitHub repo) that can do this for us:
  
```{r, results='hide'}
simdata$Population <- factor(simdata$Population, levels=seq_along(prevalences), labels=paste0('Pop_', seq_along(prevalences)))
source("autohuiwalter.R")
auto_huiwalter(simdata[,c('Population','Test1','Test2')],
               outfile='autohw.bug')
```

---
  
This generates self-contained model/data/initial values etc (ignore covse and covsp for now):
  
```{r echo=FALSE, comment=''}
cat(readLines('autohw.bug')[-(1:2)], sep='\n')
```
---
  
And can be run directly from R:
  
```{r, results='hide'}
results <- run.jags('autohw.bug')
results
```

```{r echo=FALSE}
res <- summary(results)[,c(1:3,9,11)]
res[] <- round(res, 3)
knitr::kable(res)
```

---
  
- Modifying priors must still be done directly in the model file

- The model needs to be re-generated if the data changes

- But remember that your modified priors will be reset

- There must be a single column for the population (as a factor), and all of the other columns (either factor, logical or numeric) are interpreted as being test results

- The function will soon be included in the runjags package
