---
title: "Diagnostic meta-analysis"
author: "Paolo Eusebi"
date: "16/09/2021"

theme: metropolis
aspectratio: 43
colortheme: seahorse

output:
  beamer_presentation: 
      slide_level: 2
      
params:
  presentation: yes
---

```{r setup, message=F, echo=F, warning=F, render=F}
if(params$presentation) options(width=60)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
library(mada)
library(readxl)
```

## Recap

- Important points from previous sessions

# Session 6a: Meta-Analysis of Diagnostic Test Accuracy Studies: Perfect Reference Test

## DTA-MA: perfect reference test

- There is an increasing interest in meta-analyzing data from diagnostic accuracy studies

- The data from the primary studies are summarized in a 2-by-2 cross-tabulation of the dichotomized test result against the true disease status (assuming we have a perfect reference test)

```{r, echo=FALSE, message=FALSE}
d = matrix(c("TP", "FN","FP", "TN"), nrow = 2,
           dimnames = list(c("T+","T-"),c("D+","D-")))
d %>%
  kable() %>%
  kable_styling(full_width = F, font_size = 10)
```
---

## DTA-MA: perfect reference test

- Serological tests for covid-19 from 5 studies on evaluation of ELISA assay for covid-19. (Bastos et al 2020).

```{r, echo=F}
df <- read_excel("data/bastos_serological_covid_2020.xlsx")

df %>%
  mutate(StudyID = paste("Study", 1:5)) %>%
  select(StudyID, TP, FP, FN, TN) %>%
  kable() %>%
  kable_styling(full_width = F, font_size = 10) 
```
