---
title: "Untitled"
author: "Paolo Eusebi"
date: "8/8/2021"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Four phases in architecture of diagnostic research

- **Phase I** Do test results in patients with the target disorder differ from those in normal people? Table 1 shows the architecture of this question.

- **Phase II** Are patients with certain test results more likely to have the target disorder than patients with other test results?

- **Phase III** Does the test result distinguish patients with and without the target disorder among patients in whom it is clinically reasonable to suspect that the disease is present?

- **Phase IV** Do patients who undergo this diagnostic test fare better (in their ultimate health outcomes) than similar patients who are not tested?

## Measures of diagnostic accuracy


|    | D+ | D- |
|:--:|:--:|:--:|
| T+ | TP | FP |
| T- | FN | TN |


## Measures of diagnostic accuracy

- Sensitivity = TP/D+
- Specificity = TN/D-
- PPV = TP/T+
- NPV = TN/T-

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

