---
title: "Untitled"
author: "Paolo Eusebi"
date: "8/8/2021"
output:
  beamer_presentation: default
  aspectratio: 43
  colortheme: seahorse

---

```{r setup, include=T, message=F}
knitr::opts_chunk$set(echo = F)
library(tidyverse)
library(kableExtra)
```

## Four phases in architecture of diagnostic research

- **Phase I** Do test results in patients with the target disorder differ from those in normal people? Table 1 shows the architecture of this question.

- **Phase II** Are patients with certain test results more likely to have the target disorder than patients with other test results?

- **Phase III** Does the test result distinguish patients with and without the target disorder among patients in whom it is clinically reasonable to suspect that the disease is present?

- **Phase IV** Do patients who undergo this diagnostic test fare better (in their ultimate health outcomes) than similar patients who are not tested?

## Measures of diagnostic accuracy

```{r}
df <- tibble("  " = c("T+", "T-"),
             "D+" = c("TP", "FN"),
             "D-" = c("TP", "FN"))

df %>% kable()
```


## Measures of diagnostic accuracy
```{r}
df <- tibble("  " = c("T+", "T-"),
             "D+" = c("TP", "FN"),
             "D-" = c("TP", "FN"))

df %>% kable()
```

- Sensitivity = TP/D+
- Specificity = TN/D-
- PPV = TP/T+
- NPV = TN/T-

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

